---
title: "Data Analysis Workflow Example Using MEM on cGVHD Patient Data (t-SNE, UMAP, FlowSOM, MEM, RMSD)"
author: "Copyright (c) 2016-2019 by Kirsten Diggins, Sierra Barone, and Jonathan Irish, All Rights Reserved; see EULA-MEM.text for MEM license information"
date: "March 2019"
output:
  pdf_document: 
    latex_engine: xelatex  
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

This dataset is 339 chronic graft-vs-host disease patients who have been scored 
according to 8 features representing involvement of 8 organ domains. There are 
some clear clusters of patients in this dataset, and some less clear clusters.
The goal of this exercise is to generate patient clusters through FlowSOM 
clustering following t-SNE or UMAP analysis, calculate MEM labels from each 
analysis, and then compare the MEM labels from the two analyses using the root 
mean square deviation (RMSD) in the MEM labels.

Each section of code should run in about 15 seconds or less (the t-SNE section
will take the longest). Running the entire script from start to finish (by 
choosing to Run All Chunks) should take a minute or two. 

```{r setup, include=FALSE}
# Load all necessary libraries
# If you get an error message, you will need to try re-installing packages by 
# going back to the 00_install_tools.RMD script
library(FlowSOM)
library(flowCore)
library(Biobase)
library(ggplot2)
library(MEM)
library(Rtsne)
library(uwot)
library(tidyverse)

# Load data file into R
setwd(paste(getwd(), "/datafiles/cGVHD", sep = ""))
filename <-  dir(pattern = "*.csv")
all.data = read.csv(filename)
scaled.data = all.data[, c(2:9)]
original.clusters = all.data[, c(22)]
original.tSNE = all.data[, c(23:24)]

# Set seed for reproducible results
overall_seed = 46
```

```{r t-SNE}
?Rtsne

# Run t-SNE on scaled markers
set.seed(overall_seed)
mytSNE = Rtsne(
  scaled.data,              # input scaled data
  dims = 2,        
  initial_dims = 8,         # number of initial dimensions
  perplexity = 15,          # perplexity (similar to # of nearest neighbors, 
                            # will scale with data sets, cannot be greater than
                            # the number of events minus 1 divided by 3)
  check_duplicates = FALSE, 
  max_iter = 10000          # number of iterations
)
tSNE.data = as.data.frame(mytSNE$Y)

range <- apply(apply(tSNE.data, 2, range), 2, diff)
graphical.ratio.t <- (range[1] / range[2])

# t-SNE flat dot plot (1 dot = 1 patient)
tSNE.plot <- data.frame(x = tSNE.data[, c(1)], y = tSNE.data[, c(2)])

ggplot(tSNE.plot) + coord_fixed(ratio = graphical.ratio.t) + 
  geom_point(aes(x = x, y = y), cex = 1) + 
  labs(x = "t-SNE 1", y = "t-SNE 2", title = "t-SNE on cGVHD Patient Data") + 
  theme_bw() + 
  labs(caption = "Data from Gandelman et al.,
       Hematologica 2019, 104: 189-196 \nFlow Repository: FR-FCM-ZYSU")
```
```{r FlowSOM on t-SNE}
?FlowSOM

# Run FlowSOM on the t-SNE axes
tSNE.data.mat <- as.matrix(tSNE.data)

# create flowFrame for FlowSOM input
metadata <-
  data.frame(name = dimnames(tSNE.data.mat)[[2]],
             desc = paste('t-SNE', dimnames(tSNE.data.mat)[[2]]))
metadata$range <- apply(apply(tSNE.data.mat , 2, range), 2, diff)
metadata$minRange <- apply(tSNE.data.mat , 2, min)
metadata$maxRange <- apply(tSNE.data.mat , 2, max)
tSNE.flowframe <- new("flowFrame",
                      exprs = tSNE.data.mat ,
                      parameters = AnnotatedDataFrame(metadata))

# implement the FlowSOM on t-SNE axes with target number of clusters set at 7
fSOM.t <-
  FlowSOM(
    tSNE.flowframe,        # input data has to be a flowframe
    colsToUse = c(1:2),
    xdim = 7,
    ydim = 7,
    nClus = 7,             # target number of clusters
    seed = overall_seed
  )
tSNE.FlowSOM.clusters <-
  as.matrix(fSOM.t[[2]][fSOM.t[[1]]$map$mapping[, 1]])

# plot t-SNE with FlowSOM clusters
ggplot(tSNE.plot) + coord_fixed(ratio = graphical.ratio.t) + 
  geom_point(aes(x = x, y = y, color = tSNE.FlowSOM.clusters), cex = 1.5) + 
  labs(x = "t-SNE 1", y = "t-SNE 2", title = "FlowSOM Clustering on t-SNE Axes", 
       color = "FlowSOM Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=5))) +
  labs(caption = "Data from Gandelman et al.,Hematologica 2019, 104: 189-196\nFlow Repository: FR-FCM-ZYSU")
```
```{r MEM on t-SNE/FlowSOM Clusters, warning=FALSE}
?MEM

# Run MEM on the FlowSOM clusters found using the t-SNE axes
cluster = as.numeric(as.vector((tSNE.FlowSOM.clusters)))
MEMdata = cbind(scaled.data, cluster)

MEM.values.tf = MEM(
  MEMdata,                # input data (last column should contain cluster 
                          # values)
  transform = FALSE,      
  cofactor = 0,
  choose.markers = FALSE, 
  markers = "all",
  choose.ref = FALSE,     # each cluster will be compared to all other patient
                          # clusters 
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "Mouth,GI,Eye,Joint,BSA,Sclerosis,Fascia,Liver",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.tf,          # input MEM values
  cluster.MEM = "both",
  display.thresh = 1,
  newWindow.heatmaps = FALSE,
  output.files = FALSE,
  labels = TRUE,
  only.MEMheatmap = TRUE
)

```

```{r UMAP}
# Run UMAP on scaled markers
set.seed(overall_seed)
myumap <- umap(scaled.data,       # input scaled data
               ret_model = TRUE, 
               n_threads = 1,     # to reproduce results 
               verbose = TRUE)
umap.data = as.data.frame(myumap$embedding)

range <- apply(apply(umap.data, 2, range), 2, diff)
graphical.ratio.u <- (range[1] / range[2])

# UMAP flat dot plot
UMAP.plot <- data.frame(x = umap.data[, 1], y = umap.data[, 2])

ggplot(UMAP.plot) + coord_fixed(ratio = graphical.ratio.u) +
  geom_point(aes(x = x, y = y), cex = 1) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "UMAP on cGVHD Patient Data") +
  theme_bw() + labs(caption = "Data from Gandelman et al., Hematologica 2019, 104: 189-196\nFlow Repository: FR-FCM-ZYSU")
```


```{r FlowSOM on UMAP}
# Run FlowSOM on the UMAP axes
umap.data.mat <- as.matrix(umap.data)

# create flowFrame for FlowSOM input
UMAP.metadata <-
  data.frame(name = dimnames(umap.data.mat)[[2]],
             desc = paste('UMAP', dimnames(umap.data.mat)[[2]]))
UMAP.metadata$range <-
  apply(apply(umap.data.mat, 2, range), 2, diff)
UMAP.metadata$minRange <- apply(umap.data.mat, 2, min)
UMAP.metadata$maxRange <- apply(umap.data.mat, 2, max)
umap.flowframe <- new("flowFrame",
                      exprs = umap.data.mat,
                      parameters = AnnotatedDataFrame(UMAP.metadata))

# implement the FlowSOM on UMAP axes using 7 as the target number of clusters
fSOM.u <-
  FlowSOM(
    umap.flowframe,
    colsToUse = c(1:2),
    xdim = 7,
    ydim = 7,
    nClus = 7,
    seed = overall_seed
  )
UMAP.FlowSOM.clusters <-
  as.matrix(fSOM.u[[2]][fSOM.u[[1]]$map$mapping[, 1]])

# plot FlowSOM clusters on UMAP axes
ggplot(UMAP.plot) + coord_fixed(ratio = graphical.ratio.u) + 
  geom_point(aes(x = x, y = y, color = UMAP.FlowSOM.clusters), cex = 1.5) + 
  labs(x = "UMAP 1", y = "UMAP 2", title = "FlowSOM Clustering on UMAP Axes",
       color = "FlowSOM Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=5))) +
  labs(caption = "Data from Gandelman et al., Hematologica 2019, 104: 189-196 \nFlow Repository: FR-FCM-ZYSU")
```
```{r MEM on UMAP/FlowSOM Clusters, warning = FALSE}
# Run MEM on the FlowSOM clusters found by using the UMAP axes
cluster.u = as.numeric(as.vector((UMAP.FlowSOM.clusters)))
MEMdata.u = cbind(scaled.data, cluster.u)

MEM.values.uf = MEM(
  MEMdata.u,
  transform = FALSE,
  cofactor = 0,
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "Mouth,GI,Eye,Joint,BSA,Sclerosis,Fascia,Liver",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.uf,
  cluster.MEM = "both",
  display.thresh = 1,
  newWindow.heatmaps = FALSE,
  output.files = FALSE,
  labels = TRUE,
  only.MEMheatmap = TRUE
)
```

```{r Fig.1 Data from Gandelman et al., Hematologica 2019}
# t-SNE plot with FlowSOM Clusters from Fig.1 in Gandelman et al., 
# Hematologica 2019
published.data <-
  data.frame(x = original.tSNE[, 1], y = original.tSNE[, 2])

ggplot(published.data) + coord_fixed(ratio = 0.6) + geom_point(aes(
  x = x,
  y = y,
  color = as.factor(original.clusters)
), cex = 1.5) + labs(
  x = "t-SNE 1",
  y = "t-SNE 2",
  title = "Published Clusters on t-SNE Axes",
  color = "FlowSOM Cluster"
) + theme_bw() + guides(colour = guide_legend(override.aes = list(size=5))) +
  labs(caption = "Data from Fig.1 as in Gandelman et al., Hematologica 2019, 104: 190\nFlow Repository: FR-FCM-ZYSU")
```

```{r MEM on Fig.1 Data, warning = FALSE}
# Run MEM on the FlowSOM clusters from paper
cluster = original.clusters
MEMdata.orig = cbind(scaled.data, cluster)

MEM.values.orig = MEM(
  MEMdata.orig,
  transform = FALSE,
  cofactor = 0,
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "Mouth,GI,Eye,Joint,BSA,Sclerosis,Fascia,Liver",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.orig,
  cluster.MEM = "both",
  cluster.medians = "none",
  display.thresh = 1,
  newWindow.heatmaps = FALSE,
  output.files = FALSE,
  labels = TRUE,
  only.MEMheatmap = TRUE
)
```

```{r RMSD for All Clusters}
# RMSD to compare MEM labels from the three different clusterings 
# (Fig.1 t-SNE and FlowSOM, our t-SNE and FlowSOM, and our UMAP and FlowSOM)
orig.MEM.scores = as.data.frame(MEM.values.orig[[5]])
rownames(orig.MEM.scores) = paste0(rownames(orig.MEM.scores), " (Fig.1)")
tf.MEM.scores = as.data.frame(MEM.values.tf[[5]])
rownames(tf.MEM.scores) = paste0(rownames(tf.MEM.scores), ' (t-SNE)')
uf.MEM.scores = as.data.frame(MEM.values.uf[[5]])
rownames(uf.MEM.scores) = paste0(rownames(uf.MEM.scores), ' (UMAP)')
all.MEM.values = as.matrix(rbind(tf.MEM.scores, uf.MEM.scores, orig.MEM.scores))

RMSD_vals <-
  MEM_RMSD(
    all.MEM.values,               # input all MEM values from three clusterings
    format = NULL,
    newWindow.heatmaps = FALSE,
    output.matrix = FALSE
  )
```
