---
title: "Data Analysis Workflow Example Using MEM on PBMC Cell Data (UMAP, FlowSOM, MEM, RMSD)"
author: "Copyright (c) 2016-2019 by Kirsten Diggins, Sierra Barone, and Jonathan Irish, All Rights Reserved; see EULA-MEM.text for MEM license information"
date: "March 2019"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

This data set contains 7 FCS (flow cytometry standard) files. Each FCS file 
contains a cell subset that is a well-established, phenotypically distinct 
population. This is mass cytometry data for healthy human PBMC (peripheral
blood mononuclear cells). The populations were expert gated following a t-SNE 
analysis. The first section of the code will return MEM labels for these 
expert gated populations. The goal of this exercise is to generate cell subsets 
through FlowSOM clustering following UMAP analysis, calculate MEM labels from 
this analysis, and then compare the MEM labels to the original, expert gated
populations using the root mean square deviation (RMSD) in the MEM labels.


Each section of code should run in about a minute or less. Running the entire 
script from start to finish (by choosing to Run All Chunks) should take a 
two to five minutes. 

```{r setup, include=FALSE}
# Load all libraries
# If you get an error message, you will need to try re-installing packages by 
# going back to the 00_install_tools.RMD script
library(FlowSOM)
library(flowCore)
library(Biobase)
library(ggplot2)
library(hexbin)
library(MEM)
library(tidyverse)
library(Rtsne)
library(uwot)
library(viridis)
library(ggExtra)
```

```{r MEM on Fig.1 Data, warning=FALSE}
# read files into R by setting working directory and directing R to the fcs file
# type
setwd(paste(getwd(), "/datafiles/PBMC", sep = ""))
files <-  dir(pattern = "*.fcs")
print(files)

# Run MEM on the manually gated populations (each FCS files is a pop) from paper
MEM.values.orig = MEM(
  files,                # input data (last column must contain cluster values)
  transform = TRUE,     # apply arcsinh transformation to the data
  cofactor = 15,        # set cofactor
  choose.markers = FALSE,
  markers = "12:20,22:23,25:33,35:36,38:40",  # choose columns run MEM on
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD19,CD117,CD11b,CD4,CD8,CD20,CD34,CD61,CD123,CD45RA,CD45,CD10,CD33,CD11c,CD14,CD69,CD15,CD16,CD44,CD38,CD25,CD3,IgM,HLA-DR,CD56",
  file.is.clust = TRUE, # each file is a "cluster" or cell subset
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.orig,      # input MEM values 
  cluster.MEM = "both",
  display.thresh = 3,   # display threshold for MEM scores
  newWindow.heatmaps = FALSE,
  output.files = FALSE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)

# prepare data for use in UMAP
data <- lapply(lapply(files, read.FCS), exprs)
ID <- c(1:length(data))
combined.data = as.data.frame(do.call(rbind, mapply(
  cbind, data, "File ID" = ID, SIMPLIFY = F
)))
combined.data$`File ID` <- as.numeric(combined.data$`File ID`)

# choose markers to use for downstream analysis and apply arcsinh
# transformation with a cofactor of 15
chosen.markers = combined.data[, c(12:20, 22:23, 25:33, 35:36, 38:40)]
transformed.chosen.markers <- chosen.markers %>%
  mutate_all(function(x)
    asinh(x / 15))

# set seed for reproducible results
overall_seed = 43
```

```{r UMAP}
# Run UMAP on all surface markers
set.seed(overall_seed)
myumap <-
  umap(transformed.chosen.markers,  # input scaled data
       ret_model = TRUE,
       n_threads = 1, 
       verbose = TRUE)
umap.data = as.data.frame(myumap$embedding)

range <- apply(apply(umap.data, 2, range), 2, diff)
graphical.ratio <- (range[1] / range[2])

# UMAP flat dot plot and density dot plot (1 dot = 1 cell)
UMAP.plot <- data.frame(x = umap.data[, 1], y = umap.data[, 2])

ggplot(UMAP.plot) + coord_fixed(ratio = graphical.ratio) + 
  geom_point(aes(x = x, y = y), cex = 1) + labs(x = "UMAP 1", y = "UMAP 2", 
                                                title = "UMAP on PBMC Data") + 
  theme_bw() + 
  labs(caption = "Data from Digggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63")

ggplot(UMAP.plot, aes(x = x, y = y)) + coord_fixed(ratio = graphical.ratio)  + 
  geom_bin2d(bins = 128) +
  scale_fill_viridis_c(option = "A", trans = "sqrt") + 
  scale_x_continuous(expand = c(0.1, 0)) +
  scale_y_continuous(expand = c(0.1, 0)) + labs(x = "UMAP 1", y = "UMAP 2", 
                                                title = "UMAP on PBMC Data") + 
  theme_bw() + 
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63")
```


```{r FlowSOM on UMAP}
# Run FlowSOM on the UMAP axes
umap.matrix <- as.matrix(umap.data)

# create flowFrame for FlowSOM input
UMAP.metadata <-
  data.frame(name = dimnames(umap.matrix)[[2]],
             desc = paste('UMAP', dimnames(umap.matrix)[[2]]))
UMAP.metadata$range <- apply(apply(umap.matrix, 2, range), 2, diff)
UMAP.metadata$minRange <- apply(umap.matrix, 2, min)
UMAP.metadata$maxRange <- apply(umap.matrix, 2, max)
umap.flowframe <- new("flowFrame",
                      exprs = umap.matrix,
                      parameters = AnnotatedDataFrame(UMAP.metadata))

# implement the FlowSOM on the data
fsom <-
  FlowSOM(
    umap.flowframe,      # input flowframe 
    colsToUse = c(1:2),  # columns to use 
    nClus = 15,          # target number of clusters 
    seed = overall_seed  # set seed
  )
FlowSOM.clusters <-
  as.matrix(fsom[[2]][fsom[[1]]$map$mapping[, 1]])

# plot FlowSOM clusters on UMAP axes
ggplot(UMAP.plot) + coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters),cex = 1.5) + 
  labs(x = "UMAP 1", y = "UMAP 2",title = "FlowSOM Clustering on UMAP Axes", 
       color = "FlowSOM Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=5)))+
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63")
```

```{r MEM on UMAP/FlowSOM Clusters}
# Run MEM on the FlowSOM clusters found from using UMAP axes
cluster = as.numeric(as.vector((FlowSOM.clusters)))
MEM.data = cbind(transformed.chosen.markers, cluster)

MEM.values.uf = MEM(
  MEM.data,              # input data (last column must contain cluster values)
  transform = FALSE,     # data is already scaled
  cofactor = 1,
  choose.markers = FALSE,
  markers = "all",       # use all chosen markers from previous selection
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD19,CD117,CD11b,CD4,CD8,CD20,CD34,CD61,CD123,CD45RA,CD45,CD10,CD33,CD11c,CD14,CD69,CD15,CD16,CD44,CD38,CD25,CD3,IgM,HLA-DR,CD56",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.uf,                # input MEM values
  cluster.MEM = "both",
  display.thresh = 3,           # display threshold for MEM scores
  newWindow.heatmaps = FALSE,
  output.files = FALSE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)
```


```{r RMSD for All Clusters}
# RMSD to compare labels from all populations (FlowSOM clusters vs. manually
# gated cell populations)
orig.MEM.scores = as.data.frame(MEM.values.orig[[5]])
rownames(orig.MEM.scores) = paste0(rownames(orig.MEM.scores), " (Fig.1)")
uf.MEM.scores = as.data.frame(MEM.values.uf[[5]])
rownames(uf.MEM.scores) = paste0(rownames(uf.MEM.scores), ' (UMAP)')
all.MEM.values = as.matrix(rbind(uf.MEM.scores, orig.MEM.scores))

RMSD_vals <-
  MEM_RMSD(
    all.MEM.values,                # input all MEM values from clustering and 
                                   # expert gating
    format = NULL,
    newWindow.heatmaps = FALSE,
    output.matrix = FALSE
  )
```
