---
title: "Data Analysis Workflow Example Using MEM on Polarized Myeloid Cells (UMAP, FlowSOM, MEM, RMSD)"
author: "Copyright (c) 2016-2019 by Kirsten Diggins, Sierra Barone, and Jonathan Irish, All Rights Reserved; see EULA-MEM.text for MEM license information"
date: "March 2019"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
editor_options:
  chunk_output_type: inline
---

This data set contains 8 FCS (flow cytometry standard) files. Each FCS file contains a polarized myeloid cell population from a mass cytometry experiment in which tissues were single-cell dissociated and cultured in various cytokine stimulation conditions. The populations were computationally isolated using minimum spanning tree algorithms (SPADE). The first section of the code will return MEM labels for these populations. The goal of this exercise is to generate cell subsets through FlowSOM clustering following UMAP analysis, calculate MEM labels from this analysis, and then compare the MEM labels to the original, expert gated populations using the root mean square deviation (RMSD) in the MEM labels.

Each section of code should run in about three minutes or less. Running the entire script from start to finish (by choosing to Run All Chunks) should take about five minutes. 

```{r setup, include=FALSE}
# Load all libraries
# If you get an error message, you will need to try re-installing packages by 
# going back to the 00_install_tools.RMD script
library(FlowSOM)
library(flowCore)
library(Biobase)
library(ggplot2)
library(hexbin)
library(MEM)
library(tidyverse)
library(Rtsne)
library(uwot)
library(viridis)
library(ggExtra)
library(RColorBrewer)
library(cowplot)
```

```{r MEM on Fig.1 Data, warning=FALSE}
# read files into R by setting working directory and directing R to the fcs file
# type
setwd(paste(getwd(), "/datafiles/MiMB", sep = ""))
files <-  dir(pattern = "*.fcs")
print(files)

# Run MEM on the polarized myeloid populations (each FCS files is a pop) from paper
MEM.values.orig = MEM(
  files,                # input data (last column must contain cluster values)
  transform = TRUE,     # apply arcsinh transformation to the data
  cofactor = 5,         # set cofactor
  choose.markers = FALSE,
  markers = "12:27,29:47",  # choose columns run MEM on
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD11b,CD19,Tim3,Slan,MerTK,CD64,CD36,CD164,CCR2,CD43,CD123,CD13,CD454RA,CD163,CD27,CD86,CD33,CD11c,CD14,CD32,S100A9,HLA-DR,CD206,CD16,CD120a,CCR7,CD8,CD25,CD3,CD68,CD9,CD45,PD-1,CD274,CD127",
  file.is.clust = TRUE, # each file is a "cluster" or cell subset
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
   MEM.values.orig,      # input MEM values 
   cluster.MEM = "both",
   display.thresh = 2,   # display threshold for MEM scores
   newWindow.heatmaps = FALSE,
   output.files = TRUE,
   labels = TRUE,
   only.MEMheatmap = FALSE
 )

# prepare data for use in UMAP
data <- lapply(lapply(files, read.FCS), exprs)
combined.data = as.data.frame(do.call(rbind, mapply(
  cbind, data, "File ID" = c(1:length(data)), SIMPLIFY = F
)))
colnames(combined.data)[1:length(combined.data) - 1] <- as.character(read.FCS
(files[[1]])@parameters@data[["desc"]])

# choose markers to use for downstream analysis and apply arcsinh
# transformation with a cofactor of 15
transformed.chosen.markers <- combined.data %>%
  select(contains("(v)"),-contains("bead")) %>%
  mutate_all(function(x)
    asinh(x / 5))

# set seed for reproducible results
overall_seed = 43
```

```{r UMAP}
# Run UMAP on all surface markers
set.seed(overall_seed)
myumap <-
  umap(transformed.chosen.markers,  # input scaled data
       ret_model = TRUE,
       n_threads = 1, 
       verbose = TRUE)
umap.data = as.data.frame(myumap$embedding)

range <- apply(apply(umap.data, 2, range), 2, diff)
graphical.ratio <- (range[1] / range[2])

# UMAP flat dot plot and density dot plot (1 dot = 1 cell)
UMAP.plot <- data.frame(x = umap.data[, 1], y = umap.data[, 2])

ggplot(UMAP.plot) + coord_fixed(ratio = graphical.ratio) + 
  geom_point(aes(x = x, y = y), cex = 1) + labs(x = "UMAP 1", y = "UMAP 2", 
                                                title = "UMAP on Data") + 
  theme_bw() + 
  labs(caption = "Data from Roussel et al., MiMB 2019, \nFlow Repository: FR-FCM-Z2Z8")

# density dot plot
ggplot(UMAP.plot, aes(x = x, y = y)) + 
  coord_fixed(ratio = graphical.ratio)+ geom_point(size = 0.5) +geom_density_2d_filled(bins = 29) +scale_fill_manual(values = c("NA","NA","NA","NA","NA","NA","NA",viridis::viridis(24,option = "A"))) +
labs(x = "UMAP 1", y = "UMAP 2", 
  title = "Density on UMAP Axes") + theme_bw()  + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())+theme(legend.position = "none") +ylim(1.2*min(umap.data[,2]),1.2*max(umap.data[,2]))+xlim(1.2*min(umap.data[,1]),1.2*max(umap.data[,1]))+
  labs(caption = "Data from Roussel et al., MiMB 2019, \nFlow Repository: FR-FCM-Z2Z8")
```



```{r plot heat on markers UMAP}
setwd(paste(getwd(), "/datafiles/MiMB", sep = ""))
dir.create("./output files/")

UMAP.by.marker<-as_tibble(umap.data) %>%
  bind_cols(transformed.chosen.markers)  %>%
  gather(channel, intensity, -V1, -V2) %>%
  mutate(across(channel,factor))%>%
  group_split(channel) %>%
  map(
    ~ggplot(.,aes(x= V1, y= V2, col = intensity)) +
  geom_point(size = 3) +
  scale_color_gradientn(
    colours = colorRampPalette(rev(brewer.pal(n = 11, name = "Spectral")))(5))+
  facet_grid(~ channel, labeller = function(x) label_value(x, multi_line = FALSE)) +
  coord_fixed() +
  theme_bw()+
  theme(strip.text.x = element_text(size = 20),legend.title=element_blank()))%>%
  plot_grid(plotlist = ., align = 'hv', ncol = 8)

png(paste("./output files/",strftime(Sys.time(),"%Y-%m-%d_%H%M%S"),"UMAP on transformed data.png"),height = 2000,width = 4000)
print(UMAP.by.marker)
dev.off()
```

```{r FlowSOM on UMAP}
# Run FlowSOM on the UMAP axes
umap.matrix <- as.matrix(umap.data)

# create flowFrame for FlowSOM input
UMAP.metadata <-
  data.frame(name = dimnames(umap.matrix)[[2]],
             desc = paste('UMAP', dimnames(umap.matrix)[[2]]))
UMAP.metadata$range <- apply(apply(umap.matrix, 2, range), 2, diff)
UMAP.metadata$minRange <- apply(umap.matrix, 2, min)
UMAP.metadata$maxRange <- apply(umap.matrix, 2, max)
umap.flowframe <- new("flowFrame",
                      exprs = umap.matrix,
                      parameters = AnnotatedDataFrame(UMAP.metadata))

# implement the FlowSOM on the data
fsom <-
  FlowSOM(
    umap.flowframe,      # input flowframe 
    colsToUse = c(1:2),  # columns to use 
    nClus = 8,          # target number of clusters 
    seed = overall_seed  # set seed
  )
FlowSOM.clusters <-
  GetMetaclusters(fsom)


qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, 
                           rownames(qual_col_pals)))

# plot FlowSOM clusters on UMAP axes
ggplot(UMAP.plot) + coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters),cex = 1.5) + 
  labs(x = "UMAP 1", y = "UMAP 2",title = "FlowSOM Clustering on UMAP Axes", 
       color = "FlowSOM Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=5)))+
  labs(caption = "Data from Roussel et al., MiMB 2019, \nFlow Repository: FR-FCM-Z2Z8")+ 
  scale_color_manual(values = col_vector)  
```

```{r MEM on UMAP/FlowSOM Clusters}
setwd(paste(getwd(), "/datafiles/MiMB", sep = ""))

# Run MEM on the FlowSOM clusters found from using UMAP axes
cluster = as.numeric(as.vector((FlowSOM.clusters)))
MEM.data = cbind(transformed.chosen.markers, cluster)

MEM.values.uf = MEM(
  MEM.data,                # input data (last column must contain cluster values)
  transform = FALSE,        # apply arcsinh transformation to the data
  cofactor = 0,            # set cofactor
  choose.markers = FALSE,
  markers = "all",  # choose columns run MEM on
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD11b,CD19,Tim3,Slan,MerTK,CD64,CD36,CD164,CCR2,CD43,CD123,CD13,CD454RA,CD163,CD27,CD86,CD33,CD11c,CD14,CD32,S100A9,HLA-DR,CD206,CD16,CD120a,CCR7,CD8,CD25,CD3,CD68,CD9,CD45,PD-1,CD274,CD127",
  file.is.clust = FALSE, # each file is a "cluster" or cell subset
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.uf,                # input MEM values
  cluster.MEM = "both",
  display.thresh = 2,           # display threshold for MEM scores
  newWindow.heatmaps = FALSE,
  output.files = TRUE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)
```


```{r RMSD for All Clusters}
setwd(paste(getwd(), "/datafiles/MiMB", sep = ""))

# RMSD to compare labels from all populations (FlowSOM clusters vs. manually
# gated cell populations)
orig.MEM.scores = as.data.frame(MEM.values.orig[[5]])
rownames(orig.MEM.scores) = paste0(rownames(orig.MEM.scores), " (Fig.1)")
uf.MEM.scores = as.data.frame(MEM.values.uf[[5]])
rownames(uf.MEM.scores) = paste0(rownames(uf.MEM.scores), ' (UMAP/FS)')
all.MEM.values = as.matrix(rbind(uf.MEM.scores, orig.MEM.scores))

RMSD_vals <-
  MEM_RMSD(
    all.MEM.values,                # input all MEM values from clustering and 
                                   # published populations
    format = NULL,
    newWindow.heatmaps = FALSE,
    output.matrix = TRUE
  )
```
